When does vapor pressure deficit drive or reduce evapotranspiration?

### 2017-11-27 doc/paper/vpd_et_paper.tex
hopefully changed all d to vpd, much easier to udnerstand now

\section{Methods}
n
The Penman-Monteith equation (hereafter PM) estimates ET as a function of atmospheric and land-surface variables:
\begin{linenomath*}
  \begin{equation}
      ET = \frac{\Delta R + g_a \rho_a c_p D_{s}}{\Delta + \gamma(1 + \frac{g_a}{g_s})},
  \end{equation}
\end{linenomath*}

@@

\begin{linenomath*}
  \begin{equation}
  g_{l-s} = g_0 + 1.6 \left(1 + \frac{g_1}{\sqrt{D_{s}}}\right) \frac{A}{c_s}
  \end{equation}
\end{linenomath*}

@@

\begin{linenomath*}
  \label{medlyn}
  \begin{equation}
  g_s = \text{LAI} \frac{R \,T}{P} \left( g_0 + 1.6 \left(1 + \frac{g_1}{\sqrt{D_{s}}}\right) \frac{A}{c_s}\right)
  \end{equation}
\end{linenomath*}

@@

\begin{linenomath*}
  \begin{equation}
    \label{uwue}
uWUE = \frac{GPP \cdot \sqrt{D}}{ET}
  \end{equation}
\end{linenomath*}
is relatively constant across time and space (within plant functional type). If, following \citet{Lin_2015}, we approximate $g_0$ as $0$, we can use uWUE to remove $A$ from $g_s$ in a way that makes PM analytically tractable:

\begin{linenomath*}
  \begin{equation}
  g_s = \frac{R \, T}{P} 1.6 \left(1 + \frac{g_1}{\sqrt{D_{s}}}\right) \frac{uWUE \; ET}{c_s \; \sqrt{D}}
  \end{equation}
\end{linenomath*}

@@

\begin{linenomath*}
  \begin{equation}
    ET = \frac{\Delta R + \frac{g_a\; P}{T} \left( \frac{ c_p D_{s}}{R_{air}} -  \frac{\gamma c_s \sqrt{D} }{ R* \; 1.6\; \text{ uWUE } (1 + \frac{g_1}{\sqrt{D}})} \right) }{ \Delta + \gamma}
    \label{et}
  \end{equation}
\end{linenomath*}

@@

\begin{linenomath*}
  \begin{equation}
    ET = \frac{\Delta R + \frac{g_a\; P}{T} \left( \frac{ c_p D_{s}}{R_{air}} -  \frac{\gamma c_s \sqrt{D} }{ R* \; 1.6\; \sigma \; \text{ uWUE } (1 + \frac{g_1}{\sqrt{D}})} \right) }{ \Delta + \gamma}
    \label{et_sigma}
  \end{equation}
\end{linenomath*}

@@

\begin{linenomath*}
  \begin{equation}
\sigma = - \frac{g_a \gamma c_s \sqrt{D_s} L_v P }{ \left(\text{ ET } ( \Delta + \gamma) - \Delta R - g_a \rho_a c_p D_{s}\right) 1.6 \; R\; T\; \text{ uWUE } (1 + \frac{g_1}{\sqrt{D_s}})}
    \label{et_sigma}
  \end{equation}
\end{linenomath*}

@@

\begin{linenomath*}
  \begin{equation}
    \frac{\partial \;  ET}{\partial \; D} = \frac{2\; g_a \; P}{T(\Delta + \gamma)}   \left(\frac{ c_p}{R_{air}} -  \frac{\gamma c_s }{1.6 \; R*\; \sigma \; \text{ uWUE }} \left( \frac{2 g_1 + \sqrt{D}}{2 (g_1 + \sqrt{D})^2}\right) \right)
    \label{d_et}
  \end{equation}
\end{linenomath*}

With Equation \ref{d_et} we have provided an analytical franework for ecosystem reponse to atmospheric demand perturbations. There are a few subtelties to taking the derivative in Equation \ref{d_et}: $\Delta$ ($\frac{d e_{s}}{d T}$) and $D$ are functionally related, so while taking the derivative we evaluate $\frac{\partial \; ET}{\partial \; D} = \frac{\partial \; ET} {\partial \; e_s} \frac{\partial \; e_s}{\partial \; D} \Big|_{\text{RH fixed}} + \frac{\partial \; ET}{partial \; RH} \frac{\partial \; RH}{\partial \; D} \Big|_{\text{$e_s$ fixed}}$. $RH$ and $e_s$ are assumed to be approximately orthogonal.

The $D$ dependence in Equation \ref{d_et} is a little opaque. However, mean $D$ is 1062 Pa, so $\sqrt{D}$ is 32.6 Pa$^{1/2}$, which is much less than $g_1$ (with the exception of ENF; Table \ref{pft}). So a series expansion in the limit $\frac{\sqrt{D}}{g_1} \to 0$ gives an approximation which makes the functional form more clear: \explain{Should I even include the series expansion?}

\begin{linenomath*}
  \begin{equation}
    \frac{\partial \;  ET}{\partial \; D} \approx \frac{g_a \; P}{T(\Delta + \gamma)}   \left(\frac{ c_p}{R_{air}} -  \frac{\gamma c_s }{1.6 \; R*\; \sigma \; \text{ uWUE }} \left( \frac{1}{g_{1}} - \frac{3 \sqrt{D}}{2 g_{1}^{2}} + \frac{2 \sqrt{D}^{2}}{g_{1}^{3}} - \frac{5 \sqrt{D}^{3}}{2 g_{1}^{4}} + \mathcal{O}\left(\left(\frac{\sqrt{D}}{g_1}\right)^{4}\right) \right) \right)
    \label{d_et_approx}
  \end{equation}
\end{linenomath*}

@@

\begin{table}
\caption{Definition of symbols and variables}
\centering
\begin{tabular}{l c c}
\hline
 Variable & Description & Units  \\
\hline
$e_s$  & saturation vapor pressure & Pa  \\ 
$T$  & temperature  & K \\
$\Delta$  & $\frac{\partial e_s}{\partial T}$ & Pa K$^{-1}$ \\
$R$  & net radiation at land surface minus ground heat flux & W m$^{-2}$   \\
  $g_a$  & aerodynamic conductance & m s$^{-1}$  \\
  $\rho_a$  & air density & kg m$^{-3}$  \\
  $c_p$  & specific heat capacity of air at constant pressure & J K$^{-1}$ kg$^{-1}$ \\
  $D$  & VPD & Pa  \\
  $\gamma$  & psychrometric constant & Pa K$^{-1}$   \\
  $g_s$  & stomatal conductance & m s$^{-1}$  \\
  $g_{l-s}$  & leaf-scale stomatal conductance & mol m$^{-2}$ s$^{-1}$  \\
  $R*$ & universal gas constant & J mol$^{-1}$ K$^{-1}$ \\
  $LAI$ & leaf area index & -\\
  $\sigma$ & uncertainty parameter & -\\
  $c_s$ & CO$_2$ concentration & $\mu$ mol CO$_2$ mol$^{-1}$ air\\
\hline
\multicolumn{2}{l}{$^{a}$Footnote text here.}
\end{tabular}
\end{table}

@@

An additional concern is that the $\sigma$ term may in fact be some function of $D$, in which case the dependence would need to be accounted for when taking the derivative. Figure \ref{lai_vpd_fig} plots the joint distribution of $\sigma$ and VPD, and shows that $\sigma$ is very weakly a function of VPD. Given this weak dependence, we argue that Equation \ref{d_et} is a valid approximation for ET response to $D$.

% below show histogram of site specific spearmannr? think would be useful 
\begin{figure}[h]
\centering
\includegraphics[width=20pc]{./fig03.png}
\caption{The joint distribution of $D$ and $\sigma$. $\sigma$ has only a weak dependence on $D$. ***This plot could proabably benefit from a box plot of site specific correlations, because some sites do have stronger depednence than others. Note also Figs 3 and 2 can probably be combined because this figure shoes $\sigma$'s histogram.***}
\label{lai_vpd_fig}
\end{figure}

Before calculating the sensitivity of ET to VPD, it is useful to consider the functional form of Equation \ref{d_et}. There are three terms: a scaling term for the full expression we will call Term 1 ($\frac{g_a \; P}{T(\Delta + \gamma)}$), a relatively constant offset we will call Term 2 ($\frac{c_p}{R_{air}}$), and a variable term we will call Term 3 ($\frac{\gamma c_s }{1.6 \; R\; \text{ uWUE }} \left( \frac{2 g_1 + \sqrt{D}}{2 (g_1 + \sqrt{D})^2}\right)$). All variables are positive, so the relative magnitude between Term 2 and Term 3 will determine the sign of the derivative, while Term 1 will scale the expression larger or smaller.

Term 2 minus Term 3's role in determining the sign of the sensitivty of ET to VPD makes it crucuial for answering our question ``When does VPD drive or reduce ET?'' Exploring these terms more, $c_s$ and $\gamma$ variability is relatively less than $\sigma$ and $D$ variability, so variability within PFT will be solely determined by $\sigma$ and $D$. If we fix uncertainty $\sigma$ at PFT averages, then Term 2 minus Term 3 is just a function of $D$. We can further dertmine a $D_{crit}$ where $\frac{\partial \; ET}{\partial \; D} = 0$:

\begin{linenomath*}
  \begin{equation}
D_{crit} = \frac{R_{air}}{4 c_p} \left( \frac{\gamma c_s}{1.6\; R \; \overline{\sigma} uWUE} + \sqrt{\frac{\gamma c_s}{1.6\; R \; \overline{\sigma} uWUE}\left( \frac{\gamma c_s}{1.6\; R \; \overline{\sigma} uWUE} + 8 g_1 \frac{c_p}{R_{air}}\right)} - 4 g_1 \frac{c_p}{R_{air}} \right)
\label{vpd_min_et}
  \end{equation}
\end{linenomath*}

Values of $D_{crit}$ as a function of PFT are shown in Table \ref{vpd_crit}. For any values of $D$ less than $D_{crit}$, $\frac{\partial \; ET}{\partial \; D}$ will be negative, and for values of $D$ greater than $D_{crit}$, $\frac{\partial \; ET}{\partial \; D}$ will be positive.

\begin{table}
  \label{vpd_crit}
\caption{Values of $D_{crit}$, where $\frac{\partial \; ET}{\partial \; D} = 0$, evaluated at PFT average values for $R_{air}$, $\sigma$, $\gamma$, and $c_s$. For reference, these values are also provided. For values of $D$ less than $D_{crit}$, $\frac{\partial \; ET}{\partial \; D}$ will be negative, and for values of $D$ greater than $D_{crit}$, $\frac{\partial \; ET}{\partial \; D}$ will be positive.}
\centering
\begin{tabular}{l c c c c c c}
  \hline
  PFT & $R_{air}$ & $c_s$ (ppm) & $\gamma$ & $\overline{\sigma}$ & $\overline{\sigma} \cdot uWUE$ & \textbf{$D_{crit}$ (Pa)} \\
  \hline
  CRO &  288.680920 & 372.567691& 65.351523& 0.684394&       2.602873&  \textbf{133.165438} \\
  CSH &   289.067152& 381.593622& 67.613172& 0.997224&       2.175278& \textbf{4439.564212} \\
  DBF &   288.624437& 377.449849& 63.421812& 0.881061&       2.746393&  \textbf{888.773243} \\
  ENF &  288.183849& 377.676463& 61.559242& 1.217892&       4.015362&  \textbf{978.084845} \\
  GRA &  288.425651& 377.264645& 61.598768& 0.850869&       2.281074& \textbf{1141.630778} \\
\hline
\multicolumn{2}{l}{$^{a}$Footnote text here.}
\end{tabular}
\end{table}

Figure \ref{term3} provides greater description for how (Term 2 - Term 3) varies with $D$, as a function of PFT. Equation \ref{d_et_approx} aids interpretation of Figure \ref{term3}. Larger $uWUE\cdot\overline{\sigma}$, and $g1$ shift the leading-order constant term ($\frac{1}{g_1}$) towards smaller values, and (Term 2 - Term 3) towards positive values. $uWUE$ and $g1$ are both water-use effiency type constants. Higher values corresponde to plants that are more willing to spend water on primary production and less evolved to conserve water. Figure \ref{term3} confirms our physical intuition: CROs are the least water conservative so have the smallest constant portion of Term 3, while CSH are the most water conservative and have the largest constant portion of Term 3. For the VPD-dependent terms in Equation \ref{d_et_approx}, differences in $g1$ between PFTs exert a greater influence than difference $uWUE$, as the power of $g1$ increases. Increasing $uWUE$ and $g1$ decreases the VPD-dependence, but $g1$ has the bigger effect due to its increasing powers. ENF ($g1 = 74.31$) has by far the largest VPD dependence of response, while CRO ($g1 = 183.1$) has the smallest VPD dependence.

@@

Figure \ref{term3}b shows the location of the minima of ET, as a function of $\sigma$ and $D$. For any $\sigma$ or VPD less (more) than these curves, Term 2 - Term 3 will be negative (positive). It is clear that the portion of VPD observations below/above these curves will be a strong function of $\sigma$. However, we can see some general trends. For CSH, $\frac{\partial \; ET}{\partial \; D}$ should be negative for the vast majority of observed $\sigma$ and VPD. The fraciton of positive $\frac{\partial \; ET}{\partial \; D}$ appears to be more even for ENF, GRA, and DBF, and we might expect a greater frequency of positive $\frac{\partial \; ET}{\partial \; D}$ for CRO.

While the above discussion of the sign of $\frac{\partial \; ET}{\partial \; D}$ is important to answer our research question, the magnitude of $\frac{\partial \; ET}{\partial \; D}$ will also impact statistics of $\frac{\partial \; ET}{\partial \; D}$ and the importance of $D$ variablility for $ET$ variability. So we will more closely examine the scaling Term 1: $\frac{P}{T} \propto \rho$, so this should vary little relative to aerodynamic conductance and $\Delta$. $\gamma$ should also be relatively constant, so Term 1 should be primarily a function of aerodynamic conductance and temperature (through the function $\Delta$). This makes sense, as aerodynamic conductance represents how efficently response at the surface is communicated to the atmosphere. As it increases, any plant response will be communicated more strongly to the atmosphere (and vice-versa).

@@

%%%%% replace below with idea:
\subsection{Bulk statistics of $\frac{\partial \; ET}{\partial \; D}$}

Table 3 confirms our expectations for PFT behavior of $\frac{\partial \; ET}{\partial \; D}$. For all PFTs except for CRO, average $\frac{\partial \; ET}{\partial \; D}$ is less than zero. However, $\frac{\partial \; ET}{\partial \; D}$ evaluated at the average of all variables (e.g. $\sigma$, $T$, $c_s$, $D$) is only negative for CSH and GRA. And, DBF in addition to CRO experiences $\frac{\partial \; ET}{\partial \; D}$ < 0 less than half the time. These observations highlight the effect of the nonlinear function in Figure \ref{term3}: $\frac{\partial \; ET}{\partial \; D}$ has a much steeper slope when the function is negative, and is thus more likely to be large.

The units of $\frac{\partial \; ET}{\partial \; D}$ make it difficult to interpret if $D$ is even a meaningful contributor to ET's variability. To better understand $D$'s contribution, we normalize $\frac{\partial \; ET}{\partial \; D}$ with $D$'s standard deviation to define a (linearized) relative change in ET for variations in $D$ . $D$'s contribution to ET's variability ranges between 30 - 40 W m$^{-2}$ for all PFTs except for CSH, which is about 100 W m$^{-2}$. Another meaningful comparison is to $\frac{\partial \; ET}{\partial \; R} \cdot std(R)$, as net radiation is generally the driver of ET (cite joe berry here). For all PFTs except for CSH $D$ contributes between 30 - 40 \% of $R$'s contribution to variability. For CSH the portion is much larger, about 88 \%. $D$'s variability is certiantly a primary contributor to $ET$'s variability.

So far, idealized plots and statistics have illuminated the form of $\frac{\partial \; ET}{\partial \; D}$ and how it varies with PFT. Large mean $\sigma$ and uWUE shifts CRO and DBF towards positive $\frac{\partial \; ET}{\partial \; D}$. However, the strongly nonlinear function of $\frac{\partial \; ET}{\partial \; D}$ at $\frac{\partial \; ET}{\partial \; D} < 0$ pushes $\overline{\frac{\partial \; ET}{\partial \; D}}$ negative for DBF (it does not do this for CRO because of CRO's high $g1$). ENF's low $g1$ value increases the dependence of $\frac{\partial \; ET}{\partial \; D}$ on $D$, and makes the function more strongly nonlinear. This has the side effect of pushing $\overline{\frac{\partial \; ET}{\partial \; D}}$ negative further than other PFTs for a given fraction $\frac{\partial \; ET}{\partial \; D} < 0$ and magnitude $\frac{\partial \; ET}{\partial \; D}(\overline{T,\ldots,D})$. GRA shows the opposite behavior; a relatively high $g1$  makes the function more linear, decreasing the magnitude of $-\overline{\frac{\partial \; ET}{\partial \; D}}$ for a given  [large] fraction $\frac{\partial \; ET}{\partial \; D} < 0$ and negative $\frac{\partial \; ET}{\partial \; D}(\overline{T,\ldots,D})$ (although aerodynamic conductance and Term 1 also probably have a role in this). Finally, low $uWUE$ of CSH pushes to toward by far the lowest values $\frac{\partial \; ET}{\partial \; D}$ (Figure \ref{term3}). Variability in $D$ accounts for the largest about of $ET$ variability for CSH. For the other PFTs, $D$ contributes less to $ET$ variability, but still represents about 15-20 \% of $R$'s contributions to ET variability.

>>>>

\section{Methods}
n
The Penman-Monteith equation (hereafter PM) estimates ET as a function of atmospheric and land-surface variables:
\begin{linenomath*}
  \begin{equation}
      ET = \frac{\Delta R + g_a \rho_a c_p VPD}{\Delta + \gamma(1 + \frac{g_a}{g_s})},
  \end{equation}
\end{linenomath*}

@@

\begin{linenomath*}
  \begin{equation}
  g_{l-s} = g_0 + 1.6 \left(1 + \frac{g_1}{\sqrt{VPD}}\right) \frac{A}{c_s}
  \end{equation}
\end{linenomath*}

@@

\begin{linenomath*}
  \label{medlyn}
  \begin{equation}
  g_s = \text{LAI} \frac{R \,T}{P} \left( g_0 + 1.6 \left(1 + \frac{g_1}{\sqrt{VPD}}\right) \frac{A}{c_s}\right)
  \end{equation}
\end{linenomath*}

@@

\begin{linenomath*}
  \begin{equation}
    \label{uwue}
uWUE = \frac{GPP \cdot \sqrt{VPD}{ET}
  \end{equation}
\end{linenomath*}
is relatively constant across time and space (within plant functional type). If, following \citet{Lin_2015}, we approximate $g_0$ as $0$, we can use uWUE to remove $A$ from $g_s$ in a way that makes PM analytically tractable:

\begin{linenomath*}
  \begin{equation}
  g_s = \frac{R \, T}{P} 1.6 \left(1 + \frac{g_1}{\sqrt{VPD}}\right) \frac{uWUE \; ET}{c_s \; \sqrt{VPD}}
  \end{equation}
\end{linenomath*}

@@

\begin{linenomath*}
  \begin{equation}
    ET = \frac{\Delta R + \frac{g_a\; P}{T} \left( \frac{ c_p VPD}{R_{air}} -  \frac{\gamma c_s \sqrt{VPD} }{ R* \; 1.6\; \text{ uWUE } (1 + \frac{g_1}{\sqrt{VPD}})} \right) }{ \Delta + \gamma}
    \label{et}
  \end{equation}
\end{linenomath*}

@@

\begin{linenomath*}
  \begin{equation}
    ET = \frac{\Delta R + \frac{g_a\; P}{T} \left( \frac{ c_p VPD}{R_{air}} -  \frac{\gamma c_s \sqrt{VPD} }{ R* \; 1.6\; \sigma \; \text{ uWUE } (1 + \frac{g_1}{\sqrt{VPD}})} \right) }{ \Delta + \gamma}
    \label{et_sigma}
  \end{equation}
\end{linenomath*}

@@

\begin{linenomath*}
  \begin{equation}
\sigma = - \frac{g_a \gamma c_s \sqrt{VPD} L_v P }{ \left(\text{ ET } ( \Delta + \gamma) - \Delta R - g_a \rho_a c_p VPD\right) 1.6 \; R\; T\; \text{ uWUE } (1 + \frac{g_1}{\sqrt{VPD}})}
    \label{et_sigma}
  \end{equation}
\end{linenomath*}

@@

\begin{linenomath*}
  \begin{equation}
    \frac{\partial \;  ET}{\partial \; VPD} = \frac{2\; g_a \; P}{T(\Delta + \gamma)}   \left(\frac{ c_p}{R_{air}} -  \frac{\gamma c_s }{1.6 \; R*\; \sigma \; \text{ uWUE }} \left( \frac{2 g_1 + \sqrt{VPD}}{2 (g_1 + \sqrt{VPD})^2}\right) \right)
    \label{d_et}
  \end{equation}
\end{linenomath*}

With Equation \ref{d_et} we have provided an analytical franework for ecosystem reponse to atmospheric demand perturbations. There are a few subtelties to taking the derivative in Equation \ref{d_et}: $\Delta$ ($\frac{d e_{s}}{d T}$) and $VPD$ are functionally related, so while taking the derivative we evaluate $\frac{\partial \; ET}{\partial \; VPD} = \frac{\partial \; ET} {\partial \; e_s} \frac{\partial \; e_s}{\partial \; VPD} \Big|_{\text{RH fixed}} + \frac{\partial \; ET}{partial \; RH} \frac{\partial \; RH}{\partial \; VPD} \Big|_{\text{$e_s$ fixed}}$. $RH$ and $e_s$ are assumed to be approximately orthogonal.

The $VPD$ dependence in Equation \ref{d_et} is a little opaque. However, mean $VPD$ is 1062 Pa, so $\sqrt{VPD}$ is 32.6 Pa$^{1/2}$, which is much less than $g_1$ (with the exception of ENF; Table \ref{pft}). So a series expansion in the limit $\frac{\sqrt{VPD}}{g_1} \to 0$ gives an approximation which makes the functional form more clear: \explain{Should I even include the series expansion?}

\begin{linenomath*}
  \begin{equation}
    \frac{\partial \;  ET}{\partial \; VPD} \approx \frac{g_a \; P}{T(\Delta + \gamma)}   \left(\frac{ c_p}{R_{air}} -  \frac{\gamma c_s }{1.6 \; R*\; \sigma \; \text{ uWUE }} \left( \frac{1}{g_{1}} - \frac{3 \sqrt{VPD}}{2 g_{1}^{2}} + \frac{2 \sqrt{VPD}^{2}}{g_{1}^{3}} - \frac{5 \sqrt{VPD}^{3}}{2 g_{1}^{4}} + \mathcal{O}\left(\left(\frac{\sqrt{VPD}}{g_1}\right)^{4}\right) \right) \right)
    \label{d_et_approx}
  \end{equation}
\end{linenomath*}

@@

\begin{table}
\caption{Definition of symbols and variables}
\centering
\begin{tabular}{l c c}
\hline
 Variable & Description & Units  \\
\hline
$e_s$  & saturation vapor pressure & Pa  \\ 
$T$  & temperature  & K \\
$\Delta$  & $\frac{\partial e_s}{\partial T}$ & Pa K$^{-1}$ \\
$R$  & net radiation at land surface minus ground heat flux & W m$^{-2}$   \\
  $g_a$  & aerodynamic conductance & m s$^{-1}$  \\
  $\rho_a$  & air density & kg m$^{-3}$  \\
  $c_p$  & specific heat capacity of air at constant pressure & J K$^{-1}$ kg$^{-1}$ \\
  $VPD$  & VPD & Pa  \\
  $\gamma$  & psychrometric constant & Pa K$^{-1}$   \\
  $g_s$  & stomatal conductance & m s$^{-1}$  \\
  $g_{l-s}$  & leaf-scale stomatal conductance & mol m$^{-2}$ s$^{-1}$  \\
  $R*$ & universal gas constant & J mol$^{-1}$ K$^{-1}$ \\
  $LAI$ & leaf area index & -\\
  $\sigma$ & uncertainty parameter & -\\
  $c_s$ & CO$_2$ concentration & $\mu$ mol CO$_2$ mol$^{-1}$ air\\
\hline
\multicolumn{2}{l}{$^{a}$Footnote text here.}
\end{tabular}
\end{table}

@@

An additional concern is that the $\sigma$ term may in fact be some function of $VPD$, in which case the dependence would need to be accounted for when taking the derivative. Figure \ref{lai_vpd_fig} plots the joint distribution of $\sigma$ and VPD, and shows that $\sigma$ is very weakly a function of VPD. Given this weak dependence, we argue that Equation \ref{d_et} is a valid approximation for ET response to $VPD$.

% below show histogram of site specific spearmannr? think would be useful 
\begin{figure}[h]
\centering
\includegraphics[width=20pc]{./fig03.png}
\caption{The joint distribution of $VPD$ and $\sigma$. $\sigma$ has only a weak dependence on $VPD$. ***This plot could proabably benefit from a box plot of site specific correlations, because some sites do have stronger depednence than others. Note also Figs 3 and 2 can probably be combined because this figure shoes $\sigma$'s histogram.***}
\label{lai_vpd_fig}
\end{figure}

Before calculating the sensitivity of ET to VPD, it is useful to consider the functional form of Equation \ref{d_et}. There are three terms: a scaling term for the full expression we will call Term 1 ($\frac{g_a \; P}{T(\Delta + \gamma)}$), a relatively constant offset we will call Term 2 ($\frac{c_p}{R_{air}}$), and a variable term we will call Term 3 ($\frac{\gamma c_s }{1.6 \; R\; \text{ uWUE }} \left( \frac{2 g_1 + \sqrt{VPD}}{2 (g_1 + \sqrt{VPD})^2}\right)$). All variables are positive, so the relative magnitude between Term 2 and Term 3 will determine the sign of the derivative, while Term 1 will scale the expression larger or smaller.

Term 2 minus Term 3's role in determining the sign of the sensitivty of ET to VPD makes it crucuial for answering our question ``When does VPD drive or reduce ET?'' Exploring these terms more, $c_s$ and $\gamma$ variability is relatively less than $\sigma$ and $VPD$ variability, so variability within PFT will be solely determined by $\sigma$ and $VPD$. If we fix uncertainty $\sigma$ at PFT averages, then Term 2 minus Term 3 is just a function of $VPD$. We can further dertmine a $VPD_{crit}$ where $\frac{\partial \; ET}{\partial \; VPD} = 0$:

\begin{linenomath*}
  \begin{equation}
VPD_{crit} = \frac{R_{air}}{4 c_p} \left( \frac{\gamma c_s}{1.6\; R \; \overline{\sigma} uWUE} + \sqrt{\frac{\gamma c_s}{1.6\; R \; \overline{\sigma} uWUE}\left( \frac{\gamma c_s}{1.6\; R \; \overline{\sigma} uWUE} + 8 g_1 \frac{c_p}{R_{air}}\right)} - 4 g_1 \frac{c_p}{R_{air}} \right)
\label{vpd_min_et}
  \end{equation}
\end{linenomath*}

Values of $VPD_{crit}$ as a function of PFT are shown in Table \ref{vpd_crit}. For any values of $VPD$ less than $VPD_{crit}$, $\frac{\partial \; ET}{\partial \; VPD}$ will be negative, and for values of $VPD$ greater than $VPD_{crit}$, $\frac{\partial \; ET}{\partial \; VPD}$ will be positive.

\begin{table}
  \label{vpd_crit}
\caption{Values of $VPD_{crit}$, where $\frac{\partial \; ET}{\partial \; VPD} = 0$, evaluated at PFT average values for $R_{air}$, $\sigma$, $\gamma$, and $c_s$. For reference, these values are also provided. For values of $VPD$ less than $VPD_{crit}$, $\frac{\partial \; ET}{\partial \; VPD}$ will be negative, and for values of $VPD$ greater than $VPD_{crit}$, $\frac{\partial \; ET}{\partial \; VPD}$ will be positive.}
\centering
\begin{tabular}{l c c c c c c}
  \hline
  PFT & $R_{air}$ & $c_s$ (ppm) & $\gamma$ & $\overline{\sigma}$ & $\overline{\sigma} \cdot uWUE$ & \textbf{$VPD_{crit}$ (Pa)} \\
  \hline
  CRO &  288.680920 & 372.567691& 65.351523& 0.684394&       2.602873&  \textbf{133.165438} \\
  CSH &   289.067152& 381.593622& 67.613172& 0.997224&       2.175278& \textbf{4439.564212} \\
  DBF &   288.624437& 377.449849& 63.421812& 0.881061&       2.746393&  \textbf{888.773243} \\
  ENF &  288.183849& 377.676463& 61.559242& 1.217892&       4.015362&  \textbf{978.084845} \\
  GRA &  288.425651& 377.264645& 61.598768& 0.850869&       2.281074& \textbf{1141.630778} \\
\hline
\multicolumn{2}{l}{$^{a}$Footnote text here.}
\end{tabular}
\end{table}

Figure \ref{term3} provides greater description for how (Term 2 - Term 3) varies with $VPD$, as a function of PFT. Equation \ref{d_et_approx} aids interpretation of Figure \ref{term3}. Larger $uWUE\cdot\overline{\sigma}$, and $g1$ shift the leading-order constant term ($\frac{1}{g_1}$) towards smaller values, and (Term 2 - Term 3) towards positive values. $uWUE$ and $g1$ are both water-use effiency type constants. Higher values corresponde to plants that are more willing to spend water on primary production and less evolved to conserve water. Figure \ref{term3} confirms our physical intuition: CROs are the least water conservative so have the smallest constant portion of Term 3, while CSH are the most water conservative and have the largest constant portion of Term 3. For the VPD-dependent terms in Equation \ref{d_et_approx}, differences in $g1$ between PFTs exert a greater influence than difference $uWUE$, as the power of $g1$ increases. Increasing $uWUE$ and $g1$ decreases the VPD-dependence, but $g1$ has the bigger effect due to its increasing powers. ENF ($g1 = 74.31$) has by far the largest VPD dependence of response, while CRO ($g1 = 183.1$) has the smallest VPD dependence.

@@

Figure \ref{term3}b shows the location of the minima of ET, as a function of $\sigma$ and $VPD$. For any $\sigma$ or VPD less (more) than these curves, Term 2 - Term 3 will be negative (positive). It is clear that the portion of VPD observations below/above these curves will be a strong function of $\sigma$. However, we can see some general trends. For CSH, $\frac{\partial \; ET}{\partial \; VPD}$ should be negative for the vast majority of observed $\sigma$ and VPD. The fraciton of positive $\frac{\partial \; ET}{\partial \; VPD}$ appears to be more even for ENF, GRA, and DBF, and we might expect a greater frequency of positive $\frac{\partial \; ET}{\partial \; VPD}$ for CRO.

While the above discussion of the sign of $\frac{\partial \; ET}{\partial \; VPD}$ is important to answer our research question, the magnitude of $\frac{\partial \; ET}{\partial \; VPD}$ will also impact statistics of $\frac{\partial \; ET}{\partial \; VPD}$ and the importance of $VPD$ variablility for $ET$ variability. So we will more closely examine the scaling Term 1: $\frac{P}{T} \propto \rho$, so this should vary little relative to aerodynamic conductance and $\Delta$. $\gamma$ should also be relatively constant, so Term 1 should be primarily a function of aerodynamic conductance and temperature (through the function $\Delta$). This makes sense, as aerodynamic conductance represents how efficently response at the surface is communicated to the atmosphere. As it increases, any plant response will be communicated more strongly to the atmosphere (and vice-versa).

@@

%%%%% replace below with idea:
\subsection{Bulk statistics of $\frac{\partial \; ET}{\partial \; VPD}$}

Table 3 confirms our expectations for PFT behavior of $\frac{\partial \; ET}{\partial \; VPD}$. For all PFTs except for CRO, average $\frac{\partial \; ET}{\partial \; VPD}$ is less than zero. However, $\frac{\partial \; ET}{\partial \; VPD}$ evaluated at the average of all variables (e.g. $\sigma$, $T$, $c_s$, $VPD$) is only negative for CSH and GRA. And, DBF in addition to CRO experiences $\frac{\partial \; ET}{\partial \; VPD}$ < 0 less than half the time. These observations highlight the effect of the nonlinear function in Figure \ref{term3}: $\frac{\partial \; ET}{\partial \; VPD}$ has a much steeper slope when the function is negative, and is thus more likely to be large.

The units of $\frac{\partial \; ET}{\partial \; VPD}$ make it difficult to interpret if $VPD$ is even a meaningful contributor to ET's variability. To better understand $VPD$'s contribution, we normalize $\frac{\partial \; ET}{\partial \; VPD}$ with $VPD$'s standard deviation to define a (linearized) relative change in ET for variations in $VPD$ . $VPD$'s contribution to ET's variability ranges between 30 - 40 W m$^{-2}$ for all PFTs except for CSH, which is about 100 W m$^{-2}$. Another meaningful comparison is to $\frac{\partial \; ET}{\partial \; R} \cdot std(R)$, as net radiation is generally the driver of ET (cite joe berry here). For all PFTs except for CSH $VPD$ contributes between 30 - 40 \% of $R$'s contribution to variability. For CSH the portion is much larger, about 88 \%. $VPD$'s variability is certiantly a primary contributor to $ET$'s variability.

So far, idealized plots and statistics have illuminated the form of $\frac{\partial \; ET}{\partial \; VPD}$ and how it varies with PFT. Large mean $\sigma$ and uWUE shifts CRO and DBF towards positive $\frac{\partial \; ET}{\partial \; VPD}$. However, the strongly nonlinear function of $\frac{\partial \; ET}{\partial \; VPD}$ at $\frac{\partial \; ET}{\partial \; VPD} < 0$ pushes $\overline{\frac{\partial \; ET}{\partial \; VPD}}$ negative for DBF (it does not do this for CRO because of CRO's high $g1$). ENF's low $g1$ value increases the dependence of $\frac{\partial \; ET}{\partial \; VPD}$ on $VPD$, and makes the function more strongly nonlinear. This has the side effect of pushing $\overline{\frac{\partial \; ET}{\partial \; VPD}}$ negative further than other PFTs for a given fraction $\frac{\partial \; ET}{\partial \; VPD} < 0$ and magnitude $\frac{\partial \; ET}{\partial \; VPD}(\overline{T,\ldots,VPD})$. GRA shows the opposite behavior; a relatively high $g1$  makes the function more linear, decreasing the magnitude of $-\overline{\frac{\partial \; ET}{\partial \; VPD}}$ for a given  [large] fraction $\frac{\partial \; ET}{\partial \; VPD} < 0$ and negative $\frac{\partial \; ET}{\partial \; VPD}(\overline{T,\ldots,VPD})$ (although aerodynamic conductance and Term 1 also probably have a role in this). Finally, low $uWUE$ of CSH pushes to toward by far the lowest values $\frac{\partial \; ET}{\partial \; VPD}$ (Figure \ref{term3}). Variability in $VPD$ accounts for the largest about of $ET$ variability for CSH. For the other PFTs, $VPD$ contributes less to $ET$ variability, but still represents about 15-20 \% of $R$'s contributions to ET variability.
